Notes on Validated Model Counting \\ Version of \today

### 2022-05-11 doc/mc-notes.tex
Fixed mistake in Blum derivation
Updated to reflect new CRAT format

{\bf Example 3}: Let $\ring$ be a finite field with $z = |\ring| \geq 2n$,
and let $\hashset$ be the set of functions
mapping elements of $X$ to elements of $\ring$.  For
two distinct functions $f_1$ and $f_2$ and a randomly chosen $\fhash
\in \hashset$, the probability that $h(f_1) \not = h(f_2)$ will be at least
$(1-1/z)^n \geq (1-1/2n)^n >  1/2$.  Therefore, these functions can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

@@

The CRAT clausal proof framework provides a means for creating a
checkable proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to a separable schema.  Once this
equivalence has been established, the schema can form the basis
for computations enabled by the
this representation, including trusted model counting

The CRAT format draws its inspiration from the LRAT format for Boolean formulas and the
QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
\item As with LRAT, a clause can be added as long as either 1) it is
  blocked, or 2) it satisfies the RAT property with respect to a supplied sequence
  of earlier antecedent clauses.  However, blocked clauses can only be added when defining a schema operation.
\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and $\pite$.
\begin{itemize}
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ or $\pite$ operation, the dependency sets of its arguments must
  be disjoint.
\item When an extension variable is introduced based on the $\por$
  operation, the step must cite earlier steps providing a RAT proof
  that the two arguments are mutually exclusive.
\item Boolean complement is provided implicitly by allowing the
  arguments of the extension operations to be literals and not just
  variables.
\end{itemize}
\item A CRAT proof must show that the schema is logically equivalent
  to the input formula, not just that they are equisatisfiable.
  Therefore, each deletion step must also be shown to be equivalence
  preserving, either because the clause is blocked or it follows from
  remaining clauses by the RAT property.
\item Unlike QRAT, it need not support universal quantification.
\end{itemize}

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
    $C$ & {\tt ab} & $L^{+}$ {\tt 0} & ${\tt -}C^{*}$  {\tt 0}  & Add blocked clause \\
    $C$ & {\tt ar} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RAT clause \\
     & {\tt db} & $C$             & ${\tt -}C^{+}$  {\tt 0} & Delete blocked clause \\
     & {\tt dr} & $C$             & $C^{+}$  {\tt 0} & Delete RAT clause \\
    \midrule
        & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
        & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
%%        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
  \end{tabular}
  }
\end{table}

@@

The first set of proof rules are similar to those in other clausal
proofs.  Our syntax optionally allows input clauses to be listed with
a rule of type {\tt i}.  Clauses can be added via a blocked-clause addition (command {\tt ab})
or a RAT addition (command {\tt ar}).
As described below, however, blocked clauses can only be added
to define $\pand$ and $\por$
%%, and $\pite$
operations.  The hints portion
of a blocked-clause addition lists all earlier clauses containing the
negated version of the pivot literal, with the clause IDs negated.
The hints portion of a RAT addition must contain a sequence of clause
IDs such that the added clause is RAT with respect to these clauses.
Similarly, a clause can be deleted if it blocked with respect to one
of its literals (command {\tt db}), or because it is RAT (command {\tt
  dr}) The hints portion of a RAT deletion operation must be an
ordered list of clause IDs justifying the deletion.

The second set of proof rules is unique to the CRAT format.  Each of
these indicates the addition of an extension variable.  For each case,
the rule must be followed by a sequence of blocked-clause additions
providing the defining clauses for the extension variable.

A product rule of the form ${\tt p}\;v\;l_1\;l_2$ indicates that $v$
will represent the product $l_1 \pand l_2$.  The blocked clause
additions must encode the formula $v \leftrightarrow (l_1 \land l_2)$.
Literals $l_1$ and $l_2$ must have disjoint dependency sets.

A sum rule of the form ${\tt s}\;v\;l_1\;l_2$ indicates that $v$ will
represent the disjunction $l_1 \por l_2$.  The blocked clause
additions must encode the formula $v \leftrightarrow (l_1 \lor l_2)$.
The rule also contains a sequence of clause IDs such that the clause
$\obar{l}_1 \lor \obar{l}_2$ is RAT with respect to the sequence.

%% An if-then-else rule of the form ${\tt ite}\;v\;l_1\;l_2\;l_3$
%% indicates that $v$ will represent $\pite(l_1, l_2, l_3)$.  The blocked
%% clause additions must encode the formula
%% $v \leftrightarrow \ite((l_1,l_2, l_3)$.
%% The dependency set for $l_1$ must be disjoint from those
%% of $l_2$ and $l_3$.

@@

A CRAT proof follows the same general form as a QRAT dual proof~\cite{bryant:cade:2021}---one
that ensures that each clause addition and each clause deletion
preserves equivalence.  Starting with the set of input clauses, it
produces a sequence of steps that both add and delete clauses.  Each
addition must be truth preserving and each deletion must be falsehood
preserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause
consisting of some variable or its complement.  Except for trivial
cases, the final literal will be an extension variable or its
complement.  That literal will indicate the root of the schema as the
(possibly negated) output of a schema operation.  Working from that
root backward, the schema can be extracted from the CRAT file.

@@

\begin{figure}
  \centering{
  \begin{tabular}{llllll}
    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
\midrule
    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause}\\
            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand x_3$ \\
    {\tt 2} & {\tt ab}  & {\tt 4  2 -3 0}  & {\tt 0}    & & Defining clauses for $p_4$ \\ 
    {\tt 3} & {\tt ab}  & {\tt -4 -2 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
    {\tt 4} & {\tt ab}  & {\tt -4 3 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
            & {\tt s}   & {\tt 5 2 4}   & {\tt 3 0}  & & Declare $s_5 = x_2 \por p_4$ \\
    {\tt 5} & {\tt ab}  & {\tt -5 2 4 0}  & {\tt 0}    & & Defining clauses for $s_5$ \\ 
    {\tt 6} & {\tt ab}  & {\tt  5 -2 0}    & {\tt -5 0} &  \\  %%Defining clause for $s_5$ \\ 
    {\tt 7} & {\tt ab}  & {\tt  5 -4 0}    & {\tt -5 0} & &  \\  %%Defining clause for $s_5$ \\ 
            & {\tt p}   & {\tt 6 -1 5}  &            & & Declare $p_6 = \obar{x}_1 \pand s_5$ \\
    {\tt 8} & {\tt ab}  & {\tt 6  1 -5 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
    {\tt 9} & {\tt ab}  & {\tt -6 -1 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
    {\tt 10} & {\tt ab}  & {\tt -6 5 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
             & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
    {\tt 11} & {\tt ab}  & {\tt -7 1 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
    {\tt 12} & {\tt ab}  & {\tt  7 -1 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
    {\tt 13} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
    {\tt 14} & {\tt ar}  & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
             & {\tt dr}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
  \end{tabular}
  }
  \caption{CRAT Proof \#1 for Formula $x_1 \lor x_2 \lor x_3$}
  \label{fig:c3:crat}
\end{figure}

@@

Figure~\ref{fig:c3:crat} shows an annotated version of the CRAT proof
for this example.  Clause \#1
is the input clause, and
clauses \#2--\#13 are the defining clauses for the four operations.
Each of the two sum operations lists one of the earlier defining
clauses as a proof that its arguments are mutually exclusive.
Proof clause \#14 adds the unit clause for the extension variable $s_7$.
We refer to the literal representing formula $\phi_1$ as $t_1$,
and we therefore have $t_1 = s_7$.
The unit clause indicates that extension variable $s_7$ will evaluate to
$\tautology$  for any assigment that satisfies the formula.  We
can write this as $\phi_1 \turnstile t_1$.  The
deletion step at the end turns this around, showing that
$t_1 \turnstile \phi_1$,
and therefore the input clause can be deleted, This
completes a proof that $t_1$ is logically
equivalent to the input formula.

@@

\begin{figure}
  \centering{
  \begin{tabular}{llllll}
    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
\midrule
    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause $C_1$}\\

    {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 3} & {\tt ab}  & {\tt 4 2 3 0} & {\tt 0}    & & Defining clauses for $p_4$ \\
    {\tt 4} & {\tt ab}  & {\tt -4 -2 0} & {\tt -2 0}    & & \\
    {\tt 5} & {\tt ab}  & {\tt -4 -3 0} & {\tt -2 0}    & & \\
            & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 6} & {\tt ab}  & {\tt 5 1 -4 0} & {\tt 0}    & & Defining clauses for $p_5$ \\
    {\tt 7} & {\tt ab}  & {\tt -5 -1 0} & {\tt -6 0}    & & \\
    {\tt 8} & {\tt ab}  & {\tt -5 4 0} & {\tt -6 0}    & & \\
    {\tt 9} & {\tt ar}  & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
            & {\tt dr}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
            & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
    {\tt 10} & {\tt ab}  & {\tt 6 -1 2 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
    {\tt 11} & {\tt ab}  & {\tt -6 1 0}    & {\tt -10 0} & & \\
    {\tt 12} & {\tt ab}  & {\tt -6 -2 0}    & {\tt -10 0} & & \\ 
    {\tt 13} & {\tt ar} & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
            & {\tt dr}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
            & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
    {\tt 14} & {\tt ab}  & {\tt -7 5 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
    {\tt 15} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -14 0} &  \\  
    {\tt 16} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -14 0} & &  \\
    {\tt 17} & {\tt ar}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
             & {\tt dr}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
             & {\tt dr}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
  \end{tabular}
}  
  \caption{CRAT Proof \#2A for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$}
  \label{fig:p2-bdd:crat}
\end{figure}

@@

Figure~\ref{fig:p2-bdd:schema} shows a schema representing the formula,
and Figure~\ref{fig:p2-bdd:crat} shows the associated CRAT proof.  This
proof was generated via a bottom-up strategy, such as would be created
using BDDs.  It creates schematic representations of the input
clauses and then forms their conjunction.  In the proof, unit
clauses are generated for the representations of the input clauses,
and then the input clauses are deleted.  These intermediate unit
clauses are used to justify a unit clause for the final root, and then
they are deleted.

>>>>

{\bf Example 3}: Let $\ring$ be a finite field with $z = |\ring| \geq 2n$,
and let $\hashset$ be the set of functions
mapping elements of $X$ to elements of $\ring$.  For
two distinct functions $f_1$ and $f_2$ and a randomly chosen $\fhash
\in \hashset$, the probability that $h(f_1) \not = h(f_2)$ will be at least
$\left(1-\frac{1}{z}\right)^n \geq \left(1-\frac{1}{2n}\right)^n >  1/2$.  Therefore, these functions can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

@@

The CRAT clausal proof framework provides a means to express a
checkable proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to a separable schema.  Once this
equivalence has been established, the schema can form the basis
for computations enabled by the
representation, including trusted model counting

The CRAT format draws its inspiration from the LRAT format for Boolean formulas and the
QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
  \item
  In addition to explicit clause additions and deletions, the proof contains declarations of $\pand$ and $\por$ operations.
\begin{itemize}
\item 
  These declarations implicitly add extension variables and their defining clauses to the proof.
  This is the only means for generating extension variables or adding blocked clauses to the proof.
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.  The resulting extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.  This sequence can either be provided explicitly or inferred by the proof checker.
\item Boolean complement is provided implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\end{itemize}
\item
  Clauses can also be added when they satisfy the RUP
  property, with respect to a sequence of existing clauses.  This sequence can
  be either supplied explicitly or inferred by the proof checker.
\item
  Deleting clauses requires proving that the resulting formula is not weaker.
  \begin{itemize}
    \item For an input clause or a clause declared by RUP addition, its
      deletion must be accompanied by a sequence of remaining clauses
      providing a RUP proof of the clause.  This sequence can either
      be provided explicitly or inferred by the proof checker.
    \item The clauses defining a $\pand$ or $\por$ operation are
      implicitly deleted by deleting the operation.  This can only be
      done when the only undeleted clauses containing
      references to the associated extension variable are those
      implicitly defined when the operation was declared.
      This implies that there can be no undeleted operations having
      the operation result as an argument.
  \end{itemize}
\end{itemize}

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
    $C$ & {\tt a} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RUP clause.  Explicit hint \\
    $C$ & {\tt a} & $L^{*}$ {\tt 0} &  {\tt * 0} & Add RUP clause.  Inferred hint \\
     & {\tt d} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause.  Explicit hint \\
     & {\tt d} & $C$             & {\tt * 0} & Delete RUP clause.  Inferred hint \\
    \midrule
    $C$    & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation.  Explicit hint \\ 
    $C$    & {\tt s} & $V \; L \; L$    & {\tt * 0}  & Declare $\por$ operation.  Inferred hint \\ 
%%        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
  \end{tabular}
  }
\end{table}

@@

The first set of proof rules are similar to those in other clausal
proofs.  Our syntax optionally allows input clauses to be listed with
a rule of type {\tt i}.  Clauses can be added via RUP addition
(command {\tt a}), with the sequence of antecedent clauses (the
``hint'') either provided explicitly or to be inferred (indicated by the character `{\tt *}') by the proof
checker.  Similarly for clause deletion (command {\tt d}).

The declaration of a $\pand$ operation has the form:
\begin{center}
\begin{tabular}{ccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$l_1$} & \makebox[5mm]{$l_2$} \\
\end{tabular}
\end{center}
where $i$ is a new clause ID, $v$ is a positive integer that does not
correspond to any previous variable, and $l_1$ and $l_2$ are signed
integers representing literals of existing variables.
This declaration implicitly causes three clauses to be added to the proof:
\begin{center}
\begin{tabular}{cccc}
\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
  $i$ & $v$ & $-l_1$ & $-l_2$ \\
  $i\!+\!1$ & $-v$ & $l_1$ & \\
  $i\!+\!2$ & $-v$ & $l_2$ & \\
\end{tabular}
\end{center}
The dependency sets for the arguments represented by literals $l_1$ and $l_2$ must be disjoint.

The declaration of a $\por$ operation has the form:
\begin{center}
\begin{tabular}{ccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$l_1$} & \makebox[5mm]{$l_2$} 
\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
\end{tabular}
\end{center}
where $i$ is a new clause ID, $v$ is a positive integer that does
not correspond to any previous variable, and $l_1$ and $l_2$ are
signed integers representing literals of existing variables.  Hint $H$
can consist either of the single character \texttt{*}, or it can be a
sequence of clause IDs.

This declaration implicitly causes three clauses to be added to the proof:
\begin{center}
\begin{tabular}{cccc}
\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
  $i$ & $-v$ & $l_1$ & $l_2$ \\
  $i\!+\!1$ & $v$ & $-l_1$ \\
  $i\!+\!2$ & $v$ & $-l_2$ \\
\end{tabular}
\end{center}
The explicit or implied hints must provide a RUP proof of the clause $-l_1 \lor -l_2$.

@@

A CRAT proof follows the same general form as a QRAT dual
proof~\cite{bryant:cade:2021}---one that ensures that each clause
addition and each clause deletion preserves equivalence.  With CRAT,
however, clauses are defined both explicitly and implicitly.  Starting
with the set of input clauses, the proof consists of a sequence of steps that
both add and delete clauses.  Each addition must be truth preserving
and each deletion must be falsehood preserving.  At the end, all input
clauses must have been deleted, and among the remaining clauses there
must be only a single unit clause consisting of some variable or its
complement.  Except for trivial cases, the final literal will be an
extension variable or its complement.  That literal will indicate the
root of the schema as the (possibly negated) output of a schema
operation.  Working from that root backward, the schema can be
extracted from the CRAT file.

@@

\begin{figure}
  A).  CRAT File Contents
  \begin{center}
  \begin{tabular}{lcllll}
    \multicolumn{4}{c}{File line} & & \multicolumn{1}{c}{Explanation} \\
\midrule
    \makebox[5mm][l]{\tt 1} & \makebox[7mm]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause}\\
    {\tt 2}        & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand x_3$ \\
    {\tt 5}        & {\tt s}   & {\tt 5 2 4}   & {\tt 3 0}  & & Declare $s_5 = x_2 \por p_4$ \\
    {\tt 8}        & {\tt p}   & {\tt 6 -1 5}  &            & & Declare $p_6 = \obar{x}_1 \pand s_5$ \\
    {\tt 11}         & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
    {\tt 14} & {\tt a}  & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
             & {\tt d}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
  \end{tabular}
  \end{center}
B). Proof Clauses
    \begin{center}
  \begin{tabular}{lcllll}
    \multicolumn{4}{c}{Clause} & & \multicolumn{1}{c}{Explanation} \\
\midrule
    \makebox[5mm][l]{\tt 1} & \makebox[7mm]{}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause}\\
    {\tt 2} &     & {\tt 4  2 -3 0}  &     & & Defining clauses for $p_4$ \\ 
    {\tt 3} &     & {\tt -4 -2 0}    & & & \\ %%Defining clause for $p_4$ \\ 
    {\tt 4} &     & {\tt -4 3 0}    &  & & \\ %%Defining clause for $p_4$ \\ 
            &     & {\tt -2 -4 0}   & {\tt 3 0}  & &  Mutual exclusion proof for $s_5$ \\
    {\tt 5} &     & {\tt -5 2 4 0}  &     & & Defining clauses for $s_5$ \\ 
    {\tt 6} &     & {\tt  5 -2 0}    &  &  \\  %%Defining clause for $s_5$ \\ 
    {\tt 7} &     & {\tt  5 -4 0}    &  & &  \\  %%Defining clause for $s_5$ \\ 
    {\tt 8} &     & {\tt 6  1 -5 0}  &    & & Defining clauses for $p_6$ \\ 
    {\tt 9} &     & {\tt -6 -1 0}    &  & &  \\  %%Defining clause for $p_6$ \\ 
    {\tt 10} &     & {\tt -6 5 0}    &  & &  \\  %%Defining clause for $p_6$ \\ 
            &     & {\tt -1 -6 0}   & {\tt 9 0}  & &  Mutual exclusion proof for $s_7$ \\
    {\tt 11} &     & {\tt -7 1 6 0}  &  & & Defining clauses for $s_7$ \\ 
    {\tt 12} &     & {\tt  7 -1 0}     & &  \\  %%Defining clause for $s_7$ \\ 
    {\tt 13} &     & {\tt  7 -6 0}     & &  \\  %%Defining clause for $s_7$ \\ 
    {\tt 14} &     & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
             & {\tt d}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
  \end{tabular}
  \end{center}
  \caption{CRAT file \#1 for formula $x_1 \lor x_2 \lor x_3$, and the resulting set of proof clauses}
  \label{fig:c3:crat}
\end{figure}

@@

Figure~\ref{fig:c3:crat}A shows an annotated version of the CRAT file
for this example, while \ref{fig:c3:crat}B shows the sequence of clauses that constitute the proof, including those defined implicitly, and those required for mutual exclusion checks.
   Clause 1
is the input clause.
Clauses 2--13 are added explcitly as the defining clauses for the four operations.
Also shown are the required mutual exclusion proofs for the two sum operations.
Proof clause 14 adds the unit clause for the extension variable $s_7$.
We refer to the literal representing formula $\phi_1$ as $t_1$,
and we therefore have $t_1 = s_7$.
The unit clause indicates that extension variable $s_7$ will evaluate to
$\tautology$  for any assigment that satisfies the formula.  We
can write this as $\phi_1 \turnstile t_1$.  The
deletion step at the end turns this around, showing that
$t_1 \turnstile \phi_1$,
and therefore the input clause can be deleted, This
completes a proof that $t_1$ is logically
equivalent to the input formula.

@@

\begin{figure}
  A).  CRAT File Contents
  \begin{center}
  \begin{tabular}{lcllll}
    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
\midrule

    \makebox[5mm][l]{\tt 1} & \makebox[7mm]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause $C_1$}\\
    {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
    {\tt 3} & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 6} & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 9} & {\tt a}  & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
            & {\tt d}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
    {\tt 10} & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
    {\tt 13} & {\tt a} & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
            & {\tt d}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
    {\tt 14} & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
    {\tt 17} & {\tt a}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
             & {\tt d}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
             & {\tt d}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
  \end{tabular}
  \end{center}  
B). Proof Clauses
  \begin{center}
  \begin{tabular}{lcllll}
    \multicolumn{4}{c}{Clause} & & \multicolumn{1}{c}{Explanation} \\
\midrule
    \makebox[5mm][l]{\tt 1} & \makebox[7mm]{}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause $C_1$}\\
    {\tt 2} &    & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
    {\tt 3} &   & {\tt 4 2 3 0} &     & & Defining clauses for $p_4$ \\
    {\tt 4} &   & {\tt -4 -2 0} &     & & \\
    {\tt 5} &   & {\tt -4 -3 0} &     & & \\

    {\tt 6} &   & {\tt 5 1 -4 0} &    & & Defining clauses for $p_5$ \\
    {\tt 7} &   & {\tt -5 -1 0} &     & & \\
    {\tt 8} &   & {\tt -5 4 0} &      & & \\

    {\tt 9} &   & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
            & {\tt d}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\

    {\tt 10} &   & {\tt 6 -1 2 0}  &    & & Defining clauses for $p_6$ \\ 
    {\tt 11} &   & {\tt -6 1 0}    &  & & \\
    {\tt 12} &   & {\tt -6 -2 0}    &  & & \\ 
    {\tt 13} &  & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
            & {\tt d}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
             &          & {\tt -5 -6 0} & {\tt 7 11 0}  & & Mutual exclusion proof for $s_7$ \\
    {\tt 14} &   & {\tt -7 5 6 0}  &     & & Defining clauses for $s_7$ \\ 
    {\tt 15} &   & {\tt  7 -5 0}    &  &  \\  
    {\tt 16} &   & {\tt  7 -6 0}    &  & &  \\
    {\tt 17} &   & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
             & {\tt d}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
             & {\tt d}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
  \end{tabular}
  \end{center}  
  \caption{CRAT file \#2A for formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$, and the resulting set of proof clauses}
  \label{fig:p2-bdd:crat}
\end{figure}

@@

Figure~\ref{fig:p2-bdd:schema} shows a schema representing the formula,
and Figure~\ref{fig:p2-bdd:crat} shows the associated CRAT file and proof clauses.  This
proof was generated via a bottom-up strategy, such as would be created
using BDDs.  It creates schematic representations of the input
clauses and then forms their conjunction.  In the proof, unit
clauses are generated for the representations of the input clauses,
and then the input clauses are deleted.  These intermediate unit
clauses are used to justify a unit clause for the final root, and then
they are deleted.
